---
title: "`loopBouquetPlot` for Brain Comparative Epigenome"
author: "Jianhong Ou"
vignette: >
  %\VignetteIndexEntry{`loopBouquetPlot` for Brain Comparative Epigenome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
  theme: simplex
  toc: true
  toc_float: true
  toc_depth: 4
  fig_caption: true
---

```{r preload, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
  library(InteractionSet)
  library(trackViewer)
  library(ChIPpeakAnno)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library(TxDb.Mmusculus.UCSC.mm10.knownGene)
  library(org.Hs.eg.db)
  library(org.Mm.eg.db)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

# Introduction
To illustrate how `loopBouquetPlot` can be used to help the comparison of different types of datasets, the brain comparative epigenome data will be plotted here.

There are several steps to repeat the plots:

1. Prepare the annotations

2. Import interactions

3. Import the ATAC-seq signals if available

4. Plot the data by `loopBouquetPlot` function.

The plot region are homolog regions of human and mouse shown in the Fig.1e of Zemke, N.R., Armand, E.J., Wang, W. et al. article.

# Load libraries
```{r loadPkg}
library(InteractionSet)
library(trackViewer)
library(ChIPpeakAnno)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
```
# Load data

```{r region, warning=FALSE}
## define the data range
mm_range <- GRanges("chr12:35216062-39306229")
hs_range <- GRanges('chr7:13552503-17745392')
## define the data file path
extdata <- system.file('extdata', 
                       'BrainData',
                       package='loopBouquetPlotDocumentation')
## the interactions were subsets of data downloaded from Supplementary Tables of 
## Zemke, N.R., Armand, E.J., Wang, W. et al. article
## Conserved and divergent gene regulatory programs of the mammalian neocortex.
## https://doi.org/10.1038/s41586-023-06819-6
human_loops <- read.delim(file.path(extdata, 'Supplementary_Table_22.subset.tsv'))
mouse_loops <- read.delim(file.path(extdata, 'Supplementary_Table_25.subset.tsv'))
## create the GInteractions objects
hs_gi <- InteractionSet::GInteractions(
  GRanges(sub('-', ':', human_loops$hg38_coord_bin1)),
  GRanges(sub('-', ':', human_loops$hg38_coord_bin2)),
  celltype=human_loops$celltype)
mm_gi <- InteractionSet::GInteractions(
  GRanges(sub('-', ':', mouse_loops$bin1)),
  GRanges(sub('-', ':', mouse_loops$bin2)),
  celltype=mouse_loops$celltype)
## prepare the annotations
hs_feature <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
mm_feature <- genes(TxDb.Mmusculus.UCSC.mm10.knownGene)
## list the ATAC-seq signals
hs_bws <- dir(file.path(extdata, 'human_m1_atac'),
              '*.bw', full.names = TRUE)
(names(hs_bws) <- sub('_ATAC_RPKM.bw', '', basename(hs_bws)))
mm_bws <- dir(file.path(extdata, 'mouse_mop_atac'),
              '*.bw', full.names = TRUE)
(names(mm_bws) <- sub('_ATAC_RPKM.bw', '', basename(mm_bws)))
```

# Plot the data
```{r plot, width = 10, height=10}
for(x in c('hs', 'mm')){
  ## get the range
  range <- get(paste0(x, "_range"))
  ## get the GInteraction object and subset the data
  gi <- subsetByOverlaps(get(paste0(x, "_gi")), range)
  ## get the range of all interactions and prepare annotation for them
  reg <- InteractionSet::regions(gi)
  reg <- reg[seqnames(reg) %in% seqnames(range)[1]]
  feature.gr <- subsetByOverlaps(get(paste0(x, "_feature")), range(reg))
  ## extract The gene symbols
  symbols <- ChIPpeakAnno::xget(
    feature.gr$gene_id,
    get(paste0('org.', toupper(substr(x, 1, 1)), substr(x, 2, 2), '.egSYMBOL')),
    output='first')
  keep <- !is.na(symbols)
  feature.gr <- feature.gr[keep]
  symbols <- symbols[keep]
  feature.gr$label <- symbols
  ## assign colors for the genes
  feature.gr$col <- sample(2:7, length(feature.gr), replace=TRUE)
  ## set the annotation type to 'gene'
  feature.gr$type <- 'gene'
  feature.gr$cex <- .5
  ## make sure score is available
  if(!'score' %in% colnames(mcols(gi))) gi$score <- 1
  ## split the interactions by celltype
  celltype <- strsplit(gi$celltype, ',')
  celltype <- lapply(celltype, function(.ele) gsub(' ', '', .ele))
  ## do loop for each celltype
  ct <- unique(unlist(celltype))
  ## get the bigwig file names
  bwsFile <- get(paste0(x, "_bws"))
  set.seed(123)
  for(i in ct){
    ## extract the interactions for the celltype
    gi.s <- gi[vapply(celltype, function(.ele) i %in% .ele, logical(1L))]
    bw <- NULL ## set default to NULL just in case there is not ATAC-seq data
    if(i %in% names(bwsFile)){
      bw <- importScore(bwsFile[i], ranges = range(reg), format = 'BigWig')
    }
    loopBouquetPlot(gi.s, range,
                    feature.gr,
                    coor_tick_unit = 5e4,
                    coor_mark_interval = 5e5,
                    lwd.gene = 2,
                    method = 1,
                    atacSig=bw)
    grid::grid.text(label=paste(x, i), x=.15, y=.9)
  }
}
```

## SessionInfo

```{r sessionInfo}
sessionInfo()
```
