---
title: "Figure 1 Vignette"
author: "Jianhong Ou"
vignette: >
  %\VignetteIndexEntry{Figure 1 Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
  theme: simplex
  toc: true
  toc_float: true
  toc_depth: 4
  fig_caption: true
---

```{r preload, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
  library(trackViewer)
  library(InteractionSet)
  library(org.Dr.eg.db)
  library(GenomicFeatures)
  library(GenomeInfoDb)
  library(GenomicRanges)
  library(RMariaDB)
  library(ChIPpeakAnno)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

# Introduction
The `loopBouquetPlot` is a new method introduced by **trackViewer** package to visualize genomic interactions along with annotation for NGS dataset such as HiC, HiChIP, PLAC-seq, ChIA-PET, and HiCAR data.

The Fig 1 is the showcase for the `loopBouquetPlot` compared to the heatmaps and links plot. Additional showcase are available at [`loopBouquetPlot` for Brain Comparative Epigenome](BrainComparativeEpigenome.html). There are several steps to repeat the `loopBouquetPlot`:

1. Prepare the annotations

2. Import the called interactions

3. Import the ATAC-seq signals if available

4. Plot the data by `loopBouquetPlot` function.

# Load libraries
```{r loadPkg}
library(trackViewer)
library(InteractionSet)
library(org.Dr.eg.db)
library(GenomicFeatures)
library(GenomeInfoDb)
library(GenomicRanges)
library(RMariaDB)
library(ChIPpeakAnno)
library(grid)
```

# Prepare annotation
```{r gene}
## prepare the TxDb object,
## or load the refseq gene annotation by library(TxDb.Drerio.UCSC.danRer11.refGene)
txdb <- GenomicFeatures::makeTxDbFromEnsembl(release=109,
                                             organism='Danio rerio')
## set the silencer coordinates
silencer <- GRanges("chr5", IRanges(21795000, 21800000))
## set data range
range <- GRanges("chr5", IRanges(21100000, 22100000))
## set the plot range
fig1PlotRange <- GRanges('chr5', IRanges(21300000, 2.2e+07))

## prepare gene annotation
genes <- GenomicFeatures::genes(txdb)
GenomeInfoDb::seqlevelsStyle(genes) <- "UCSC"
genes <- subsetByOverlaps(genes, range, ignore.strand=TRUE)
id <- genes$gene_id
eid <- ChIPpeakAnno::xget(id, org.Dr.egENSEMBL2EG, output = "first")
symbol <- id
symbol[!is.na(eid)] <- ChIPpeakAnno::xget(eid[!is.na(eid)],
                                          org.Dr.egSYMBOL,
                                          output = "first")
names(symbol) <- id

feature.gr <- subsetByOverlaps(genes, range, ignore.strand=TRUE)
feature.gr$label <- symbol[feature.gr$gene_id]
# remove genes without symbols to clean the plot
feature.gr <- feature.gr[!grepl('ENSDARG', feature.gr$label)]
## show selected gene symbols only
feature.gr$label[!feature.gr$label %in% c('tenm1', 'smarca1', 'sh2d1aa')] <- ''
feature.gr$type <- 'gene'
feature.gr$col <- 1 # set black color to uninterested genes
feature.gr$col[feature.gr$label %in% c('tenm1', 'smarca1', 'sh2d1aa')] <- 
  c(2, 4, 5) ## set the colors for the given genes
silencer$label <- 'silencer'
silencer$type <- 'cRE' ## cis regulatory element
silencer$col <- 3 ## set the color to green
feature.gr <- c(feature.gr, silencer)
```

# Subfigure __b__ and __c__

## Import data for `loopBouquetPlot`
```{r import4loopBouquetPlot}
## set the data folder, all data are available in the extdata folder of this package
extdata <- system.file('extdata', package='loopBouquetPlotDocumentation')
## list all interaction BEDPE files
(interactions <- dir(extdata,
                     "10k.2.sig3Dinteractions.bedpe",
                     full.names = TRUE))
## import the loop for `loopBouquetPlot`
loops_loopBouquet <- lapply(interactions, function(.ele){
  ## read bedpe as text file
  .loop <- read.delim(.ele, header = FALSE)
  ## create GInteractions object
  .loop <- with(.loop, InteractionSet::GInteractions(
    ## narrow the interaction region for better visualization effect
    GRanges(V1, IRanges(V2+2501, V3-2500)),
    GRanges(V4, IRanges(V5+2501, V6-2500)),
    ## set the score to log2 value to decrease the difference of signals
    score = log2(V8)
  ))
  ## subset the data with given range
  .loop <- subsetByOverlaps(.loop, range)
  ## remove the interchromosal interactions
  .loop <- .loop[seqnames(first(.loop))==seqnames(range)[1] &
                   seqnames(second(.loop))==seqnames(range)[1]]
})
## create names for the data for each sample
(names(loops_loopBouquet) <- sub(".10k.*$", " interactions",
                                 basename(interactions)))

## list all teh ATAC-seq bigwig files
(bws <- dir(extdata,
            "bigWig",
            full.names = TRUE))
## import the ATAC-seq signales
## the ATAC-seq signals are HiCAR R2 read coverage
bwsigs <- lapply(bws, trackViewer::importScore, format="BigWig", ranges=range)
## create names for the data for each sample
(names(bwsigs) <-sub('.bigWig', ' ATAC-seq', basename(bws)))
## check the order of the samples are same for ATAC-seq signals and loops
stopifnot(all(sub(' .*$', '', names(bwsigs))==
                sub(' .*$', '', names(loops_loopBouquet))))
```

## Plot the data by `loopBouquetPlot` for subfigure __b__ and __c__
```{r loopBouquetPlot}
set.seed(1) ## set seed to control the plot layout
for(i in seq_along(loops_loopBouquet)){
  trackViewer::loopBouquetPlot(loops_loopBouquet[[i]],
                               fig1PlotRange,
                               feature.gr,
                               atacSig=bwsigs[[i]],
                               lwd.backbone=2,
                               col.backbone='gray',
                               col.backbone_background = 'gray70',
                               lwd.gene = 2,
                               show_coor = TRUE,
                               coor_tick_unit = 1e4,
                               coor_mark_interval=2e5,
                               length.arrow=0.1,
                               show_edges = FALSE)
  grid::grid.text(label=names(loops_loopBouquet)[i],
            x=.15, y=.9)
}
```

# Subfigure __a__

There are several steps to repeat the heatmaps and links plot:

1. Prepare the annotations

2. Import the ATAC-seq signals

3. Import the called interactions for link tracks

4. Import the interactions for heatmap tracks

5. Plot the data by `viewTracks` function.

## Import data for subfigure __a__
```{r import4viewTracks}
## import the interactions as 'track' objects
loops <- lapply(interactions, function(.ele){
  .loop <- read.delim(.ele, header = FALSE)
  .loop <- with(.loop, GInteractions(
    GRanges(V1, IRanges(V2+2501, V3-2500)),
    GRanges(V4, IRanges(V5+2501, V6-2500)),
    score = log2(V8)
  ))
  .loop <- gi2track(.loop)
  ## set the tracktype to 'link' to plot as links
  setTrackStyleParam(.loop, "tracktype", "link")
  .loop
})
names(loops) <- sub(".10k.*$", " interactions", basename(interactions))
## get the hic files. Here we use the prepared GInteraction objects
(hics <- dir(extdata,
             "10000.ginteractions.rds",
             full.names = TRUE))
## set the sample names
(names(hics) <- sub("MAPS_signals_(.*?).10000.ginteractions.rds", "\\1 heatmap",
                    basename(hics)))
gis <- lapply(hics, function(.ele){
  ## The RDS files were generated by trackViewer function `importGInteractions`
  ## from the HIC file. Because the MAPS signal are values normalized by
  ## fitting models, we set the matrixType to 'observed' and normalization to 'NONE'
  ## range.ext are ranges with extention to include all the intrachromosal interactions
  # range.ext <- range
  # start(range.ext) <- start(range.ext) - 10000000
  # end(range.ext) <- end(range.ext) + 10000000
  # importGInteractions(.ele, ranges = range.ext,
  #                     format = "hic",
  #                     resolution = 10000,
  #                     out = "GInteractions",
  #                     normalization = "NONE",
  #                     matrixType = "observed")
  .ele <- readRDS(.ele)
})
## check the order of the samples
stopifnot(identical(sub(" .*$", "", names(gis)),
                    sub(" .*$", "", names(loops))))
## create heatmap signals
heatmaps <- mapply(gis, interactions, FUN=function(.gi, .ele){
  .loop <- read.delim(.ele, header = FALSE)
  .loop <- with(.loop, GInteractions(
    GRanges(V1, IRanges(V2+2501, V3-2500)),
    GRanges(V4, IRanges(V5+2501, V6-2500)),
    score = log2(V8)
  ))
  ## highlight the called interactions by add color the border
  .gi$border_color <- NA
  ol <- findOverlaps(.gi, .loop)
  .gi$border_color[queryHits(ol)] <- "black"
  ## create 'track' object from GInteraction object
  gi2track(.gi)
})
## create gene track object
trs <- geneTrack(id, txdb, symbol, asList = FALSE)
seqlevelsStyle(trs$dat) <- "UCSC" ## make sure the seqnames are in same style
```

## Plot the subfigure __a__ by viewTracks
```{r viewTracks,fig.width=7,fig.height=10}
## create the plot style by default algorithm
op <- optimizeStyle(
  trackList = trackList(trs, 
                        rev(bwsigs),   ## reverse the signal
                        rev(loops),    ## to keep 0 dpa samples
                        rev(heatmaps), ## in top
                        ## set the height for each track
                        heightDist = c(2, length(bwsigs),
                                       length(loops),
                                       length(heatmaps)*8)))
tL <- op$tracks
sty <- op$style
for(i in which(grepl('ATAC-seq', names(tL)))){#R2 ATAC-seq signal
  ## set the ylim. The raw range is [0,1]
  setTrackStyleParam(tL[[i]], "ylim", c(0.15, .8))
  ## set the y-axis to the right side
  setTrackYaxisParam(tL[[i]], "main", FALSE)
  ## plot the y-axis lables
  setTrackYaxisParam(tL[[i]], "label", TRUE)
  ## draw the y-axis
  setTrackYaxisParam(tL[[i]], "draw", TRUE)
}
## set color keys for the link tracks
for(i in which(grepl('interactions', names(tL)))){
  setTrackStyleParam(tL[[i]], "breaks",
                     ## set the breaks from min to max
                     c(seq(from=2, to=6.5, by=.5), 10))
  setTrackStyleParam(tL[[i]], "color",
                     c("lightblue", "yellow", "red"))
  setTrackYaxisParam(tL[[i]], "draw", TRUE)
}
## set color keys for the heatmaps
for(i in which(grepl('heatmap', names(tL)))){
  setTrackStyleParam(tL[[i]], "breaks", 
                     c(seq(from=0, to=5, by=.5), 8.5))
  setTrackStyleParam(tL[[i]], "color",
                     c("#FDF8F5", "#EA7556", "#4D0E0B"))
}
##
trackViewer::viewTracks(tL, viewerStyle = sty,
                        gr=fig1PlotRange, autoOptimizeStyle = FALSE)

## set the TADs to plot with the heatmap
tad <- GRanges("chr5",
               IRanges(c(21060000, 21220000, 21326000, 21890000, 22000000),
                       c(21130000, 21290000, 21860000, 21960000, 22090000)))
## add the annotations to the heatmap
for(i in which(grepl('heatmap', names(tL)))){
  ## add TADs as dashed line (lty=2)
  trackViewer::addInteractionAnnotation(tad, i, grid.lines,
                                        gp=gpar(col="black",
                                                lwd=1, lty=2))
  ## add the guide line for the silencer with slope -1 with lwd=3
  trackViewer::addInteractionAnnotation(-21795000, i,
                                        gp=gpar(col = "#009E73",
                                                lwd=3, lty=3))
}
```

## SessionInfo

```{r sessionInfo}
sessionInfo()
```
