---
title: "`loopBouquetPlot` example for loops caller comparison"
author: "Jianhong Ou"
bibliography: bibliography.bib
csl: nature.csl
vignette: >
  %\VignetteIndexEntry{`loopBouquetPlot` example for loops caller comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
  theme: simplex
  toc: true
  toc_float: true
  toc_depth: 4
  fig_caption: true
---

```{r setup, include=FALSE}
library(InteractionSet)
library(trackViewer)
library(geomeTriD)
library(ChIPpeakAnno)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
library(rtracklayer)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
To demonstrate the utility of `loopBouquetPlot` in comparing different loop callers for the same dataset, we will visualize the loops called by `MAPS`[@juric2019maps] and `hichipper`[@lareau2018hichipper] side by side.

```{r loadlibrary}
library(InteractionSet)
library(trackViewer)
library(geomeTriD)
library(ChIPpeakAnno)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
library(rtracklayer)
```

## DATA from MAPS paper

Download data from MAPS paper.
```{r, warning=FALSE, eval=FALSE}
url <- 'https://doi.org/10.1371/journal.pcbi.1006982.s027'
tmpf <- tempfile()
download.file(url=url, destfile = tmpf)
tmpdir <- tempdir()
unzip(tmpf, exdir = tmpdir)
d <- file.path(tmpdir, 'Supplementary data')
MAPS_files <- dir(d, 'MAPS')
(names(MAPS_files) <- sub('_combine_MAPS_call', '',
                          sub('.bedpe', '', MAPS_files)))
hichipper_files <- dir(d, 'hichipper')
(names(hichipper_files) <- sub('_combine_hichipper_converted.bedpe', 
                               '', hichipper_files))
maps_loops <- lapply(MAPS_files, function(.ele){
  .ele <- read.delim(file.path(d, .ele))
  with(.ele, GInteractions(GRanges(chr1, IRanges(start1+1, end1)),
                           GRanges(chr2, IRanges(start2+1, end2)),
                           score = -log10(fdr)))
})
hichipper_loops <- lapply(hichipper_files, function(.ele){
  .ele <- read.delim(file.path(d, .ele))
  with(.ele, GInteractions(GRanges(chr1, IRanges(start1+1, end1)),
                           GRanges(chr2, IRanges(start2+1, end2)),
                           score = -log10(fdr)))
})

mm_gi <- list(
  maps_CTCF         = maps_loops[["mESC_CTCF"]],
  hichipper_CTCF    = hichipper_loops[["mESC_CTCF"]],
  maps_H3K4me3      = maps_loops[["mESC_H3K4me3"]],
  hichipper_H3K4me3 = hichipper_loops[["mESC_H3K4me3"]]
)
```

We have already downloaded the data and saved a subset of it within this package. Now, we will load the data from this package to save time.
```{r}
## set the data folder, all data are available in the extdata folder of this package
extdata <- system.file('extdata',
                       'GSE119663',
                       package='loopBouquetPlotDocumentation')
## list all interaction BEDPE files
interactions <- dir(extdata,
                     "bedpe",
                     full.names = TRUE)
names(interactions) <- sub('.bedpe.gz',
                           '',
                           basename(interactions))
mm_gi <- lapply(interactions, import)
mm_gi <- lapply(mm_gi, function(.ele){
  GInteractions(first(.ele), second(.ele), score=mcols(.ele)$score)
})
## hic files were downloaded from processed data of GSE119663
mm_hic <- dir(extdata, 'PLACseq.rds', full.names = TRUE)
(names(mm_hic) <- sub("_PLACseq.rds", " heatmap", basename(mm_hic)))
```

Here is the helper function to prepare the annotation data.
```{r helperFun}
prepareFeatures <- function(gi, range, species){
  reg <- lapply(gi, InteractionSet::regions)
  reg <- unlist(GRangesList(reg))
  reg <- reg[seqnames(reg) %in% seqnames(range)[1]]
  feature.gr <- subsetByOverlaps(get(paste0(species, "_feature")), range(reg))
  ## extract The gene symbols
  symbols <- ChIPpeakAnno::xget(
    feature.gr$gene_id,
    org.Mm.egSYMBOL,
    output='first')
  feature.gr$symbols <- symbols
  geneType <- ChIPpeakAnno::xget(
    feature.gr$gene_id,
    org.Mm.egGENETYPE,
    output='all'
  )
  keep <- !is.na(symbols) & symbols!='NA' & grepl('protein-coding', geneType)
  feature.gr$label <- NA
  feature.gr$label[keep] <- symbols[keep]
  ## assign colors for the genes
  feature.gr$col[keep] <- sample(2:7, length(feature.gr[keep]), replace=TRUE)
  ## set the annotation type to 'gene'
  feature.gr$type <- 'gene'
  feature.gr
}
```

## Plot data by `loopBouquetPlot` for a given region

```{r plot, fig.show="hold", out.width="50%", width=14, height=14}
mm_range <- GRanges('chr10', IRanges(60600001, 61660000))
mm_feature <- genes(TxDb.Mmusculus.UCSC.mm10.knownGene)
mm_feature.gr <- prepareFeatures(mm_gi, mm_range, species='mm')

op <- par(mfrow=c(2, 2))
for(caller in c('maps', 'hichipper')){ ## loop for caller
  for(antibody in c('CTCF', 'H3K4me3')){ ## loop for antibody type
    gi.s <- subsetByOverlaps(mm_gi[[paste0(caller, '_', antibody)]], mm_range) ## get the interaction data
    if(length(gi.s)>1){
      ## plot data
      set.seed(123)
      loopBouquetPlot(gi = gi.s,
                      range = mm_range,
                      feature.gr = mm_feature.gr,
                      coor_tick_unit = 5e4,
                      coor_mark_interval = 5e5,
                      lwd.gene = 2,
                      method = 1)
      ## mark the sub-figure
      grid::grid.text(label=paste(caller, antibody), x=.15, y=.9)
    }
  }
}
par(op)
```

```{r, heatmap, height=14}
gis <- lapply(mm_hic, function(.ele){
  ## data were prepared by importGInteractions with KR normalization
  ## score are log2 transformed observed data
    # .ele <- importGInteractions(.ele, ranges = range.ext,
    #                     format = "hic",
    #                     resolution = 10000,
    #                     out = "track",
    #                     normalization = "KR",
    #                     matrixType = "observed")
    # .ele$dat$score <- log2(.ele$dat$score)
    readRDS(.ele)
  })
id <- mm_feature.gr$gene_id
symbol <- mm_feature.gr$symbols
trs <- geneTrack(id, TxDb.Mmusculus.UCSC.mm10.knownGene,
                 symbol, asList = FALSE)
trs$dat$color[!is.na(trs$dat$id)] <-
  mm_feature.gr[trs$dat$id[!is.na(trs$dat$id)]]$col
## loops
loops <- lapply(mm_gi, gi2track)
loops <- lapply(loops, function(.ele){
  .ele$dat$score <- log2(.ele$dat$score)
  setTrackStyleParam(.ele, "tracktype", "link")
  .ele
})
names(loops) <- paste(names(loops), 'interactions')
op <- optimizeStyle(
  trackList = trackList(trs, 
                        rev(loops),
                        gis,
                        ## set the height for each track
                        heightDist = c(2,
                                       length(loops),
                                       length(gis)*8)))
tL <- op$tracks
sty <- op$style
## set color keys for the link tracks
for(i in which(grepl('interactions', names(tL)))){
  setTrackStyleParam(tL[[i]], "breaks",
                     ## set the breaks from min to max
                     c(seq(from=2, to=6.5, by=.5), 10))
  setTrackStyleParam(tL[[i]], "color",
                     c("lightblue", "yellow", "red"))
  setTrackYaxisParam(tL[[i]], "draw", TRUE)
  setTrackStyleParam(tL[[i]], 'ylabpos', 'bottomleft')
}
## set color keys for the heatmaps
for(i in which(grepl('heatmap', names(tL)))){
  setTrackStyleParam(tL[[i]], "breaks", 
                     c(seq(from=0, to=6.5, by=.5), 10))
  setTrackStyleParam(tL[[i]], "color",
                     c("#FDF8F5", "#EA7556", "#4D0E0B"))
}
vp <- trackViewer::viewTracks(tL, viewerStyle = sty,
                        gr=mm_range, autoOptimizeStyle = FALSE)
addGuideLine(c(60750000, 61070000, 61162000, 61460000), col='red', vp=vp)
tads <- GRanges('chr10', IRanges(c(60260001, 60750001, 61070001, 61460001),
                                   c(60750000, 61070000, 61460000, 61735000)))
for(i in which(grepl('heatmap', names(tL)))){
  addInteractionAnnotation(tads, names(tL)[i],
                         grid.lines, gp=gpar(col="#0072B2", lwd=3, lty=3))
}
## highlight the differential interaction region among the H3K4me3 and CTCF PLAC-Seq
gi_sub <- GInteractions(
  GRanges('chr10', IRanges(c(60260001, 60750001, 61070001),
                           c(60750000, 61070000, 61460000))),
  GRanges('chr10', IRanges(c(60750001, 61070001, 61460001),
                           c(61070000, 61210000, 61735000)))
)
## make sure not olverlap with TAD curve
start(regions(gi_sub)) <- start(regions(gi_sub))+5000
end(regions(gi_sub)) <- end(regions(gi_sub))-5000
addInteractionAnnotation(gi_sub, "H3K4me3 heatmap",
                         grid.polygon,
                         gp=gpar(col="#009E73", lwd=2, lty=2, fill=NA))
```

### Plot FLAMINGO results

Here we use the raw single normalized by KR as FLAMINGO input. 
The FLAMINGO results will be ploted in the same region above.


```{r FLAMINGO, warning=FALSE, message=FALSE}
## list all FLAMINGO output files
flamingo <- dir(extdata, pattern=".*chr10.flamingo.rds", full.names = TRUE)
names(flamingo) <- 
  sub('^(.*).chr10.flamingo.rds', '\\1', basename(flamingo) )
flamingo.gr <- lapply(flamingo, readRDS)
flamingo.gr <- lapply(flamingo.gr,
                      function(.ele) subsetByOverlaps(.ele, mm_range, type = 'within'))
view3dStructure(flamingo.gr[['CTCF']],
                k=3, feature.gr=mm_feature.gr,
                length.arrow=
                  unit(abs(diff(range(flamingo.gr[['CTCF']]$x)))/20,
                       'native'))
eID <- 'flamingo.CTCF'
rglwidget(elementId = eID)
widget <- lapply(c('tick_labels', 'tick_minor', 'tick_major', 
                   'atac_signal', 'backbone', 'gene_body', 'tss_labels',
                   'gene_labels', 'cRE'), function(.ele){
                     toggleWidget(sceneId = eID, tags = .ele)
                   })
manipulateWidget::combineWidgets(list=widget)

view3dStructure(flamingo.gr[['H3K4me3']],
                k=3, feature.gr=mm_feature.gr,
                length.arrow=
                  unit(abs(diff(range(flamingo.gr[['H3K4me3']]$x)))/20,
                       'native'))
eID <- 'flamingo.H3K4me3'
rglwidget(elementId = eID)
widget <- lapply(c('tick_labels', 'tick_minor', 'tick_major', 
                   'atac_signal', 'backbone', 'gene_body', 'tss_labels',
                   'gene_labels', 'cRE'), function(.ele){
                     toggleWidget(sceneId = eID, tags = .ele)
                   })
manipulateWidget::combineWidgets(list=widget)
```

# sessionInfo
```{r}
sessionInfo()
```

